A free/open-source machine translation system for English to Kazakh

### 2015-03-07 engkaz-eamt-2015/engkaz-eamt-2015.tex
Mikel's ironing is done
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59254 72bbbca6-d526-0410-a7d9-f06f51895060

\usepackage{polyglossia}
\usepackage{todonotes}
\setdefaultlanguage[variant=australian]{english}
\usepackage{latexsym}
%\usepackage[small,bf]{caption}
\usepackage[bf]{caption}
\usepackage{xltxtra}
\usepackage{times}
\usepackage{fontspec}
\usepackage{footnote}
\usepackage{natbib}
\defaultfontfeatures{Scale=MatchLowercase,Mapping=tex-text}
\usepackage{listings}
\usepackage{microtype}

@@

\author{Aida Sundetova\\
  Kazakh National University,  \\
  050040 Almaty, \\
  {\tt sun27aida@gmail.com}\\
  \And
  \textbf{Francis M. Tyers}\\
  HSL-fakultetet\\ 
  UiT Norgga \'{a}rktala\v{s} universitehta \\
  9017 Romsa, \\
  {\tt francis.tyers@uit.no}
  \And
  Mikel L. Forcada\\
  Dept. Lleng. i Sist. Inform. \\
  Universitat d'Alacant, \\ 
  Alacant \\
  {\tt mlf@dlsi.ua.es}  
}

@@

This paper presents a shallow-transfer rule-based machine translation (MT) system from English 
(a West Germanic language) to Kazakh (a Turkic language)
The complex agglutinative morphology of Kazakh 
is in contrast with the simple morphology of English.
Turkic language morphology shows clear morphotactics (ordering of morphemes), but with 
complex phonological changes (morphophonology) to due to morpheme contact (vowel harmony, 
sonorisation, etc.) many of which are explicitly represented in writing.

@@

We set out to build a rule-based machine translation system, because for statistical machine translation 
system we need large sentence-aligned parallel corpora, which are not freely available
for the English--Kazakh pair. An English--Kazakh statistical MT system has however recently been built by Google,\footnote{\url{http://translate.google.com}}; 
other work in this language pair was performed by building a small parallel corpus and using Moses \citep{assylbekov14}. Existing 
commercial systems for English to Kazakh (Sanasoft\footnote{\url{http://www.sanasoft.kz/c/ru/node/47}} 
and Trident\footnote{\url{http://www.translate.ua/us/on-line}}) both appear to be rule-based.

@@

\begin{figure*}[htbp]
\begin{center}
 \includegraphics[width=0.8\textwidth]{architecture.pdf}
\end{center}
\caption{The pipeline architecture of the Apertium system.}
\label{fig:modules}
\vspace{-1em}
\end{figure*}

@@

has been developed since 2005, started at the Universitat d'Alacant. At first, it was initially aimed at 

@@

free/open-source GNU General Public License (GPL, versions 2 and 3).
\todo{If space allowed, it would be nice to add a running example as in the Apertium paper}

@@

\item[Part-of-speech (POS) tagger:] This module chooses one of the LF of an ambiguous SF. 

@@

 Multiword units can be translated as a single word.

@@

\item [Structural transfer:] This module uses pattern matching to identify sequences of LF (phrases or 
  segments), which need syntactical processing 

@@

It uses files with rules, which specify  the syntactic transformation such as word reorderings, 
  lexical changes such as changes in prepositions and agreement between target language lexical forms. %Transfer rules can 

@@

\item[Morphological generator:] Transforms the sequence of target-language LFs forms, produced by the structural transfer, 

@@

For Kazakh, a finite-state transducer based on two-level rules \citep{washingtonsalimzyantyers14} is used for generation. 
\item[Post-generator:] Performs some minor orthographical operations in the target language. 
\item[ Reformatter:] It places format tags back into the text so that its format is preserved.

@@

Apertium mainly uses XML-based formats for linguistic data, which include bilingual and monolingual dictionaries, structural transfer and
lexical selection rules.

@@

The English dictionary is used to determine lexical forms of each surface
forms, it contains of list of all lexical units, definition of
source-language alphabet, grammatical categories, which is used to
specify lexical forms, for example, like noun, verb, gender, etc.;
each category has paradigms describing groups of correspondences
between parts of SFs and LFs.

The Kazakh dictionary is turned into a
morphological transducer, using the Helsinki Finite
State Toolkit \citep{hfst/2011}. It uses the \texttt{lexc} formalism
for defining lexicons through word classes and subclasses, and the
\texttt{twol} formalism for morphophonological rules such as vowel
harmony, desonorization, nasalization, etc.  Depending on the
preceding vowels and consonants, the plural suffix \emph{-LAr}\footnote{Uppercase Latin letters are used for archiphonemes (actually archigraphemes) that are realised as phonemes (actually graphemes) after morphophonological rules have been applied} could
become \emph{-тар}, \emph{-тер}, \emph{-лар}, \emph{-лер}, \emph{-дар}, \emph{-лер} depending on vowel harmony: кітап+тар,
мектеп+тер, etc. 
%\todo{Aida, please typeset this example in Cyrillic; braces are not used here, but are used later; DONE  homogeneize notation in paper.}

@@

The English--Kazakh bilingual dictionary provides correspondences between English LFs and Kazakh LFs. Ambiguity is allowed: one of the LFs will be chosen by 
lexical-selection rules depending on context. This dictionary currently contains 13,135 stems \citep{sundetova13a}.

@@

This kind of rules solves the problem of part-of-speech (PoS) ambiguity using the  Constraint Grammar (CG, \cite{karlsson95}) formalism. CG rules choose one of the LFs obtained for each SF. %From English to Kazakh POS ambiguity could be only in English side, but also some choice also
%depend on Kazakh grammar.\todo{I removed an unclear sentence here}

@@

Lexical-selection rules choose one of the alternative target-language LFs corresponding to one source-language LF, as described in \cite{tyers12a}. Alternative translations are defined in bilingual dictionary by 
multiple entries for each source-language LF. For example, the adjective \emph{beautiful} could be translated as \emph{сұлу}, if 
the following noun is a person: \emph{beautiful girl} \(\to\) \emph{сұлу қыз}; or as \emph{әдемі} or  \emph{көркем}, if the following noun means 
place or \todo{or what? - mountains} \emph{beautiful mountains} \(\to\) \emph{әдемі таулар}.  Table~\ref{table:lexsel} shows some 
lexical-selection rules:\todo{but this rule is not in the table!!! - should I add it or just remove upper example??}
\begin{table*}
 \centering
 \begin{tabular}{|l|l|l|l|}
    \hline
    \textbf{SL word \(w\)} & \textbf{TL word} & \textbf{Context} & \textbf{Example} \\
    \hline
    \multirow{4}{*}{residence} & \multirow{2}{*}{мекен}      & \multirow{2}{*}{default} & I live in a residence. \\
                               &                             &                     & (Мен) мекенде өмір сүремін. \\
                               & \multirow{2}{*}{резиденция} & \multirow{2}{*}{\(w\)\_ of \_ president} & I see the residence of the president \\ 
                               &                             &                                     & (Мен) президенттің резиденциясын көремін \\
    \hline
    \multirow{4}{*}{boot} & \multirow{2}{*}{бәтеңке} & default & I bought boots. \\
                          &                          &     &  Мен бәтеңкелерді сатып алдым \\

@@

    \hline
 \end{tabular}
  \caption{Example lexical-selection rules}
  \label{table:lexsel}
\end{table*}
\todo{Yes, it is mean I see whatever - I do not understand the \emph{I see anything} example. Does it mean \emph{anything} as \emph{whatever}? I do not know how to put the \(w\) in this example}@

@@

\begin{itemize}
\item A first round of transformations (``chunker'') detects source language (SL) LF patterns and generates the 
  corresponding sequences of TL LFs grouped in chunks representing simple constituents such as noun phrases, prepositional phrases, etc. 
\item The second round (``interchunk'') reads patterns of chunks and produces a new sequence of chunks. This is the 
  module where one can attempt to perform some longer-range reordering operations, interchunk agreement, case selection, etc. 
\item The third round (``postchunk'') transfers chunk-level tags to the lexical forms they contain and whose lexical-form-level 
  tags are linked (through a referencing systems) to chunk-level tags (for instance, case determined for a noun phrase is 
  transferred to the main noun), and removes all grouping information to generate the desired sequence of TL LFs.
\end{itemize}

The structural transfer module in Apertium processes the stream of source-language lexical form -- target-language lexical 
form pairs (SL LF–TL LF pairs) and transforms it into a new sequence of TL LFs after a series of structural transfer 
operations specified in a set of rules: reordering, elimination or insertion of TL LFs, agreement, etc.

This section describes the current structural transfer in \texttt{apertium-eng-kaz}.\todo{Complete! 58998? Move to results? - moved} English--Kazakh 
chunker rules, interchunk rules and an additional clean-up stage will be described in detail in the following 3 sections.

\paragraph{Chunker:}\todo{This should have 4 numerals, but I don't know how to do this}
In the first round of structural transfer, rules segment sentences into chunks, like short noun 
phrases, adjective phrases, verb phrases and adpositional phrases (that is, prepositional phrases in English and 
postpositional phrases in Kazakh).
Chunking rules, of which there are currently 168, identify 8 kinds of chunks and translate them into equivalent Kazakh chunks, leaving some adaptations to be performed in later stages of structural transfer (for instance, the 
morphological case of noun phrases).

@@

\begin{savenotes}
\begin{table*}
  \centering
  \begin{tabular}{|l|l|l|l|}
    \hline
    \textbf{Phrase type} & \textbf{Description} & \textbf{Examples} & \textbf{Chunk tags} \\
    \hline
    \multirow{5}{*}{\texttt{NP}} & \multirow{5}{*}{Noun phrase}    & noun & number*, person, \\
                                 &                                 & determiner-noun & possessor*, case\\
                                 &                                 & numeral-noun & \\
                                 &                                 & adjective-noun & \\
                                 &                                 & determiner-adjective-noun & \\
    \hline
    \multirow{5}{*}{\texttt{VP}} & \multirow{5}{*}{Verb phrase}    & finite-verb & number, person, \\
                                 &                                 & have-participle & tense/mood, possessive*, \\
                                 &                                 & be-gerund & negation* \\
                                 &                                 & must-infinitive & \\ 
                                 &                                 & want\_to-infinitive & \\ 
    \hline
    \multirow{5}{*}{\texttt{PP}} & \multirow{5}{*}{Postpositional phrase} & preposition-noun & number, person, \\
                                 &                                        & preposition-determiner-noun & possessor*, case \\
                                 &                                        & preposition-numeral-noun & \\
                                 &                                        & preposition-adjective-noun & \\
                                 &                                        & preposition-determiner-adjective-noun & \\
    \hline
    \multirow{2}{*}{\texttt{GenP}} & \multirow{2}{*}{Genitive phrase}  & As with \texttt{PP} except & number, person, \\
                                   &                                   & with the preposition `of'. & possessor*, case \\
    \hline
    \multirow{2}{*}{\texttt{AdjP}} & \multirow{2}{*}{Adjective phrase}  & adjective & n/a \\
                                   &                                    & ``more''--adjective & \\
    \hline
    \multirow{2}{*}{\texttt{SupP}} & \multirow{2}{*}{Superlative phrase} & adjective--``est'' & possessor*, case \\
                                   &                                     & the--``most''--adjective & \\ 
    \hline
%    \texttt{Q\_mark} & Question mark  & & \\
%    \hline
  \end{tabular}
  \caption{Phrase types in the English-Kazakh chunker}
  \label{table:phrases} 
\end{table*}
\end{savenotes}
As it is shown above, the most probable sequence of world classes are built phrases and 
for English-Kazakh chunker there are used next kind of phrases:

@@

\item[Noun phrases:] Consider the following example: the chunker identifies the English sequence 
  the large book  (determiner–adjective–noun) as a noun-phrase chunk. It translates into 
  Kazakh, and assigns it four chunk-level tags: number (set to singular), person (set to 3rd), 

@@

  ending (\emph{кітабы}) later if the context were, for instance, \emph{the large book of animals}, \emph{аңдардың үлкен кітабы}), 
  and case (to be determined as it could be, for instance, accusative in \emph{I saw the large book}, \emph{Мен үлкен кітапты көрдім} ). Also, the
  noun form of the verb, for instance,  gerund form is determined as NP phrase (\emph{Her playing is good} – \emph{Оның ойыны жақсы}).\todo{Make sure all English and Kazakh examples are in italics}

@@

    \item Genitive -{N}{I}ң,\footnote{Upper case letters in braces (such as {N}) represent hypothetical archiphonemes (actually archigraphemes) that are realised as phonemes (actually graphemes) after morphophonological rules have been applied. For instance, in the genitive ending “-{N}{I}ң”, the archiphoneme {N} may be realised as т, д, or н and the archiphoneme {I} may be realised as  і  or ы depending on the previous phonological context. } which will be marked GenP:  
        [PP [P of ] [NP the beautiful garden] ] $\rightarrow$ [GenP [NP әдемі бақша] [P -ның] ] 
    \item Translating into any of the morphological cases: locative ``-{D}{A}'',\footnote{{D} can be д or т, and {A} can be е or а, depending on the 
        phonological context.} ablative ``-{D}{A}н'', etc., except the genitive ``-{N}{I}ң'' case: 

@@

\item[Verb phrases:] Translation of English verb phrases into Kazakh  are not always be straightforward, for instance: 
  % present simple and future are rendered using the same tense in Kazakh (I play $\rightarrow$ Мен ойнаймын;  I 
  % will play $\rightarrow$ Мен ойнаймын), this ambiguity problem may be solved by lexical selection, 
  % which will choose right tense by watching context; 

@@

  found in the -п participle form (\emph{ойнап}), and a suitable finite form  (\emph{жатырмын}, \emph{отырдым}) of an auxiliary 
  verb (\emph{жатыр}, \emph{отыр}) is used to express number and person agreement,\footnote{Actually, Kazakh language uses four auxiliary verbs: \emph{жатыр} `lie', 

@@

      is \emph{жатыр} `lie' for the present continuous and отыр (`sit') for the past continuous.} 
  modal verbs are translated by adding adjectives, which means \emph{керек} `necessary' or \emph{жөн} `proper', \todo{This part is hard to read. The choice of auxilliary is not justified.}
  and a form of the copula (absent in present tense); the subject receives the genitive or dative case; negative 
  verbs for present continuous and past tense are translated by adding negative word ``емес'' for present 
  perfect continuous (He has not been playing -- Ол ойнап отырған емес) and  ``жоқ'' for past tense and 
  present continuous (see examples above).

@@

\item Comparative adjective phrases  (English more + adjective, or adjective-[e]r); the Kazakh translation 
   chooses the comparative suffix ``-{I}р{A}{K}'': [AdjP  \emph{more  interesting} ]  $\rightarrow$ [AdjP \emph{қызығ\emph{ырақ}}]
\item For superlative adjective phrases  ``the most + adjective''  or ``adjective-[e]st'', translation 

@@

   [SupP  \emph{the most important}]   $\rightarrow$ [SupP \emph{ең маңызды}]

@@

modified by a genitive phrase: \emph{the most beautiful of people} $\rightarrow$ \emph{адамдардың ең әдемі\textbf{сі}}); they are treated as a noun phrase with an implied noun (``the largest [book]'').
\todo{The table of phrase types is not referred to in the text; also, the meaning of ``*'' is not explained}

@@

\begin{table*}
  \centering
  \begin{tabular}{|l|l|l|l|}
    \hline
    \multicolumn{4}{|c|}{\textbf{Affirmative forms}} \\
    \hline
    \textbf{English form} & \textbf{Example} & \textbf{Morphemes} & \textbf{Translation} \\
    \hline
    \multirow{2}{*}{Present simple} & \multirow{2}{*}{I play} & \multirow{2}{*}{ойна + <aorist> + <pers.>} & Мен ойнаймын \\
                                    &                         &                                            & I play \\
    \hline
    \multirow{2}{*}{Present continuous} & \multirow{2}{*}{I am playing} & ойна + <perf. part.>  & Мен ойнап жатырмын \\
                                    &                         &   жатыр + <pres.> + <pers.>     & I playing lie \\
    \hline
    \multirow{2}{*}{Past continuous} & \multirow{2}{*}{I was playing} & ойна + <perf. part.>  & Мен ойнап отырдым \\
                                    &                         & отыр + д{I} <past>  + <pers.>     & I playing sat \\
    \hline
    \multirow{2}{*}{Obligation} & \multirow{2}{*}{I must go} & Мен  + <gen.>  & Менің баруым керек \\
                                    &                         & бар + <ger.> +  <px1sg> + керек    & My going necessary [is] \\
    \hline
    \multirow{2}{*}{Should} & \multirow{2}{*}{I should go} & Мен  + <gen.>  & Менің Менің барғаным жөн \\
                                    &                         & бар + <past.ger.> +  <px1sg> + жөн & My was.going proper [is] \\
    \hline
    \multicolumn{4}{|c|}{\textbf{Negative forms}} \\
    \hline
    \multirow{2}{*}{Present cont.} & \multirow{2}{*}{I am not playing} & ойна + <perf. part.>  & Мен ойнап жатқан жоқпын \\
                                    &                         &   жатыр + <pres.> + <pers.>     & I playing lying not \\
    \hline
    \multirow{2}{*}{Ability} & \multirow{2}{*}{I can not play} & ойна + <impf. part.>  & Мен ойнай алмаймын \\
                                    &                         &   ал + <neg.> + <aor.> + <p1>     & I playing can.not \\
    \hline
    \multirow{3}{*}{Present perf. cont. } & \multirow{3}{*}{You have not been playing} & ойна + <perf. part.>  & Сіз ойнап отырған емессі \\
                                    &                         & +  отыр + <past>  & You playing sat not  \\
                                    &                         & + eмес <cop.neg> + <p2> & \\
    \hline

  \end{tabular}
  \caption{Examples of Verb phrase translations}
  \label{table:vps}
\end{table*}

\paragraph{Interchunk processing:}
\todo{Table~\ref{table:vps} should be typeset with examples in italics and making sure that the margins are respected}

>>>>

\usepackage{polyglossia}
\usepackage{todonotes}
\setdefaultlanguage[variant=australian]{english}
\usepackage{latexsym}
%\usepackage[small,bf]{caption}
\usepackage[bf]{caption}
\usepackage{xltxtra}
\usepackage{times}
\usepackage{fontspec}
\usepackage{footnote}
\usepackage{natbib}
\defaultfontfeatures{Scale=MatchLowercase,Mapping=tex-text}
\usepackage{listings}
\usepackage{microtype}
\newcommand{\anonymize}[1]{\phantom{#1}}
%\newcommand{\anonymize}[1]{#1}

@@

\author{\anonymize{Aida Sundetova}\\
  \anonymize{Kazakh National University, } \\
  \anonymize{050040 Almaty, Kazakhstan} \\
  \anonymize{{\tt sun27aida@gmail.com}}\\
  \And
  \anonymize{\textbf{Francis M. Tyers}}\\
  \anonymize{HSL-fakultehta}\\ 
  \anonymize{UiT Norgga \'{a}rktala\v{s} universitehta} \\
  \anonymize{N-9017 Romsa, Norway} \\
  \anonymize{{\tt francis.tyers@uit.no}}
  \And
  \anonymize{Mikel L.\ Forcada}\\
  \anonymize{Dept.\ Lleng.\ i Sist.\ Informàt.} \\
  \anonymize{Universitat d'Alacant,} \\ 
  \anonymize{E-03071 Alacant, Spain} \\
  \anonymize{{\tt mlf@dlsi.ua.es}}
}

@@

This paper presents a shallow-transfer rule-based machine translation (MT) system from English 
(a West Germanic language) to Kazakh (a Turkic language). This is a very challenging language pair for MT.
The complex agglutinative morphology of Kazakh 
is in contrast with the simple morphology of English.
Turkic language morphology shows clear morphotactics (ordering of morphemes), but with 
complex phonological changes (morphophonology) to due to morpheme contact (vowel harmony, 
sonorisation, etc.) many of which are explicitly represented in writing.

@@

We set out to build a rule-based machine translation system, because for statistical machine translation 
system we need large sentence-aligned parallel corpora, which are not freely available
for the English--Kazakh pair. An English--Kazakh statistical MT system has however recently been built by Google;\footnote{\url{http://translate.google.com}}
other work in this language pair was performed by building a small parallel corpus and using Moses \citep{assylbekov14}. Existing 
commercial systems for English to Kazakh (Sanasoft\footnote{\url{http://www.sanasoft.kz/c/ru/node/47}} 
and Trident\footnote{\url{http://www.translate.ua/us/on-line}}) both appear to be rule-based.

@@

\begin{figure*}[htbp]
\begin{center}
 \includegraphics[width=0.8\textwidth]{architecture.pdf}
\end{center}
\caption{The pipeline architecture of the Apertium system.}
\label{figure:pipeline}
\vspace{-1em}
\end{figure*}

@@

has been developed since 2005, started at the Universitat d'Alacant. Apertium Was initially aimed at 

@@

free/open-source GNU General Public License (GPL, versions 2 and 3). The Apertium pipeline (see figure~\ref{figure:pipeline}) has the following components:
%\todo{If space allowed, it would be nice to add a running example as in the Apertium paper}

@@

\item[Part-of-speech (POS) tagger:] This module chooses one of the LFs of an ambiguous SF. 

@@

 Multiword units can be translated as a single LF.

@@

\item [Structural transfer:] This module uses pattern matching to identify sequences of LFs (phrases or 
  segments), which need syntactical processing.

@@

It uses files with rules, which specify syntactic transformations such as word reorderings, 
  lexical changes such as changes in prepositions, and agreement between target language lexical forms. %Transfer rules can 

@@

\item[Morphological generator:] It transforms the sequence of target-language LFs, produced by the structural transfer, 

@@

\item[Post-generator:] Performs some minor orthographical operations in the target language.\footnote{This module is not used when the target language is Kazakh.} 
\item[ Reformatter:] It places format tags back into the text so that its format is preserved.
\end{description}

@@

Apertium uses text-based (mainly XML-based) formats for linguistic
data, which include bilingual and monolingual dictionaries, structural
transfer and lexical selection rules.

@@

Morphological dictionaries are used to define the correspondences between LFs and SFs; they contain a definition of the source-language alphabet and grammatical categories and attributes (such as noun, verb, plural, locative, etc.), a list of all lexical units, and inflection paradigms for all lexical categories; paradigms describe regularities in the correspondences
between parts of SFs and LFs.

The English dictionary comes from existing data in Apertium, such as the English---Spanish language pair.

The Kazakh dictionary \citep{washingtonsalimzyantyers14} is turned
into a finite-state morphological transducer, using the Helsinki
Finite State Toolkit \citep{hfst/2011}. The dictionary is based on
two-level rules, and uses the \texttt{lexc} formalism for defining
lexicons through word classes and subclasses, and the \texttt{twol}
formalism for morphophonological rules such as vowel harmony,
desonorization, nasalization, etc. An example of morphophonological
rule: depending on the preceding vowels and consonants, the plural
suffix \emph{-LAr}\footnote{Uppercase Latin letters are used for
  archiphonemes (actually archigraphemes) that are realised as
  phonemes (actually graphemes) after morphophonological rules have
  been applied.} for noun could become \emph{-тар}, \emph{-тер}, \emph{-лар},
\emph{-лер}, \emph{-дар}, \emph{-лер}:
кітап+тар, мектеп+тер, etc.

@@

The English--Kazakh bilingual dictionary provides correspondences between English LFs and Kazakh LFs. Ambiguity is allowed: one of the LFs will be chosen by 
lexical-selection rules depending on context (see below). This dictionary currently contains 13,135 stems \citep{sundetova13a}.

@@

The problem of part-of-speech (PoS) ambiguity is solved using the  Constraint Grammar (CG, \cite{karlsson95}) formalism: CG rules choose one of the LFs obtained for each SF. %From English to Kazakh POS ambiguity could be only in English side, but also some choice also
%depend on Kazakh grammar.\todo{I removed an unclear sentence here}

@@

Lexical-selection rules choose one of the alternative target-language LFs corresponding to one source-language LF, as described by \cite{tyers12a}. Alternative translations are defined in bilingual dictionary by 
multiple entries for each source-language LF. For example, the adjective \emph{beautiful} could be translated as \emph{сұлу}, if 
the following noun is a person: \emph{beautiful girl} \(\to\) \emph{сұлу қыз}; or as \emph{әдемі} or  \emph{көркем}, if the following noun means 
place: \emph{beautiful mountains} \(\to\) \emph{әдемі таулар}.  Table~\ref{table:lexsel} shows examples of 
lexical-selection rules:

\begin{table*}
 \centering
 \begin{tabular}{|l|l|l|l|}
    \hline
    \textbf{SL word \(w\)} & \textbf{TL word} & \textbf{Context} & \textbf{Example} \\
    \hline
    \multirow{4}{*}{residence} & \multirow{2}{*}{мекен}      & \multirow{2}{*}{default} & I live in a residence. \\
                               &                             &                     & (Мен) мекенде өмір сүремін. \\
                               & \multirow{2}{*}{резиденция} & \multirow{2}{*}{\(w\)\_ of \_ president} & I see the residence of the president \\ 
                               &                             &                                     & (Мен) президенттің резиденциясын көремін \\
    \hline
    \multirow{4}{*}{boot} & \multirow{2}{*}{бәтеңке} & default & I bought boots. \\
                          &                          &     &  Мен бәтеңкелерді сатып алдым \\

@@

    \hline
 \end{tabular}
  \caption{Example lexical-selection rules}
  \label{table:lexsel}
\end{table*}

@@

\begin{itemize}
\item A first round of transformations (``chunker'') detects source language (SL) LF patterns and generates the 
  corresponding sequences of target language (TL) LFs grouped in chunks representing simple constituents such as noun phrases, prepositional phrases, etc. 
\item The second round (``interchunk'') reads patterns of chunks and produces a new sequence of chunks. This is the 
  module where one can attempt to perform some longer-range reordering operations, interchunk agreement, case selection, etc. 
\item The third round (``postchunk'') transfers chunk-level tags to the lexical forms they contain and whose lexical-form-level 
  tags are linked (through a referencing systems) to chunk-level tags (for instance, case determined for a noun phrase is 
  transferred to the main noun), and removes all grouping information to generate the desired sequence of TL LFs.
\end{itemize}

The structural transfer module in Apertium processes the stream of source-language lexical form -- target-language lexical 
form pairs (SL LF–TL LF pairs) and transforms it into a  sequence of TL LFs after a series of structural transfer 
operations specified in a set of rules: reordering, elimination or insertion of TL LFs, agreement, etc.

This section describes the current structural transfer in \texttt{apertium-eng-kaz}.  English--Kazakh 
chunker rules, interchunk rules and an additional clean-up stage will be described in detail in the following 3 sections.

\paragraph{Chunker:}
In the first round of structural transfer, rules segment sentences into chunks, like short noun 
phrases, adjective phrases, verb phrases and adpositional phrases (that is, prepositional phrases in English and 
postpositional phrases in Kazakh).
Chunking rules, of which there are currently 168, identify 8 kinds of chunks and translate them into equivalent Kazakh chunks, leaving some adaptations to be performed in later stages of structural transfer (for instance, the 
morphological case of noun phrases).

@@

\begin{savenotes}
\begin{table*}
  \centering
  \begin{tabular}{|l|l|l|l|}
    \hline
    \textbf{Phrase type} & \textbf{Description} & \textbf{Examples} & \textbf{Kazakh chunk tags} \\
    \hline
    \multirow{5}{*}{\texttt{NP}} & \multirow{5}{*}{Noun phrase}    & noun & number, person, \\
                                 &                                 & determiner-noun & possessor, case\\
                                 &                                 & numeral-noun & \\
                                 &                                 & adjective-noun & \\
                                 &                                 & determiner-adjective-noun & \\
    \hline
    \multirow{5}{*}{\texttt{VP}} & \multirow{5}{*}{Verb phrase}    & finite-verb & number, person, \\
                                 &                                 & have-participle & tense/mood, possessive, \\
                                 &                                 & be-gerund & negation \\
                                 &                                 & must-infinitive & \\ 
                                 &                                 & want\_to-infinitive & \\ 
    \hline
    \multirow{5}{*}{\texttt{PP}} & \multirow{5}{*}{Postpositional phrase} & preposition-noun & number, person, \\
                                 &                                        & preposition-determiner-noun & possessor, case \\
                                 &                                        & preposition-numeral-noun & \\
                                 &                                        & preposition-adjective-noun & \\
                                 &                                        & preposition-determiner-adjective-noun & \\
    \hline
    \multirow{2}{*}{\texttt{GenP}} & \multirow{2}{*}{Genitive phrase}  & As with \texttt{PP} except & number, person, \\
                                   &                                   & with the preposition `of'. & possessor, case \\
    \hline
    \multirow{2}{*}{\texttt{AdjP}} & \multirow{2}{*}{Adjective phrase}  & adjective & (no tags) \\
                                   &                                    & ``more''--adjective & \\
    \hline
    \multirow{2}{*}{\texttt{SupP}} & \multirow{2}{*}{Superlative phrase} & adjective--``est'' & possessor, case \\
                                   &                                     & the--``most''--adjective & \\ 
    \hline
%    \texttt{Q\_mark} & Question mark  & & \\
%    \hline
  \end{tabular}
  \caption{Phrase types in the English--Kazakh chunker. }
  \label{table:phrases} 
\end{table*}
\end{savenotes}
As it is shown above, the most probable sequence of world classes are built phrases and 
for English-Kazakh chunker there are used next kind of phrases:

@@

\item[Noun phrases:] Consider the following example: the chunker identifies the English LF sequence for \emph{the large book}  (determiner--adjective--noun) as a noun-phrase chunk. It translates it into the Kazakh adjective--noun sequence \emph{үлкен кітап}, and assigns it four chunk-level tags: number (set to singular), person (set to 3rd), 

@@

  ending (\emph{кітабы}) later if the context were, for instance, \emph{the large book of animals}, \emph{аңдардың үлкен кіта\textbf{бы}}), 
  and case, to be determined later as it could be, for instance, accusative in \emph{I saw the large book}, \emph{Мен үлкен кітап\textbf{ты} көрдім}). Similar processing occurs for nominal forms of the verb: for instance, as in the NP \emph{her paying} in \emph{Her playing is good}, which is  \emph{Оның ойыны жақсы} (\emph{She}+\textsc{genitive} \emph{playing}+\textsc{poss-3rd} \emph{good}). 

@@

    \item Genitive \emph{-NIң},\footnote{In the genitive ending \emph{-{N}{I}ң}, the archiphoneme {N} may be realised as т, д, or н and the archiphoneme {I} may be realised as  і  or ы depending on the previous phonological context. } in which will case the phrase will be marked GenP:  
        [PP [P \emph{of} ] [NP \emph{the beautiful garden}] ] $\rightarrow$ [GenP [NP \emph{әдемі бақша}] [P \emph{-ның}] ] 
    \item Translating into any of the morphological cases: locative \emph{-{D}{A}}'',\footnote{\emph{D} can be \emph{д} or \emph{т}, and \emph{{A}} can be \emph{е} or \emph{а}, depending on the 
        phonological context.} ablative \emph{-{D}{A}н}, etc. --- all except the genitive \emph{-{N}{I}ң} case: 

@@

\item[Verb phrases:] Translation of English verb phrases into Kazakh is not always be straightforward. For instance,

@@

  found in the \emph{-п} participle form (\emph{ойна\textbf{п}}), and a suitable finite form  (\emph{жатырмын}, \emph{отырдым}) of an auxiliary 
  verb (\emph{жатыр}, \emph{отыр}) is used to express number and person agreement\footnote{Actually, Kazakh uses four auxiliary verbs: \emph{жатыр} `lie', 

@@

      is \emph{жатыр} `lie' for the present continuous and отыр (`sit') for the past continuous.}. Modal verbs are translated by adding adjectives, which means \emph{керек} `necessary' or \emph{жөн} `proper', 
  and a form of the copula (absent in present tense), with the subject receiving the genitive or dative case. Negative 
  verbs for present continuous and past tense are translated by adding the negative word \emph{емес} for present 
  perfect continuous (\emph{He has not been playing} \(\to\) \emph{Ол ойнап отырған емес}) and  \emph{жоқ} for past tense and 
  present continuous.

@@

\item Comparative adjective phrases  (English \emph{more} + adjective, or adjective-[\emph{e}]\emph{r}); the Kazakh translation 
   chooses the comparative suffix \emph{{I}р{A}{K}}: [AdjP  \emph{more  interesting} ]  $\rightarrow$ [AdjP \emph{қызығ\textbf{ырақ}}]
\item For superlative adjective phrases  (\emph{the most} + adjective  or adjective-[\emph{e}]\emph{st}, translation 

@@

   [SupP  \emph{the most important}]   $\rightarrow$ [SupP \emph{ең маңызды}].

@@

modified by a genitive phrase: \emph{the most beautiful of people} $\rightarrow$ \emph{адамдардың ең әдемі\textbf{сі}}); they are treated as a noun phrase with an implied noun (\emph{the largest} [\emph{book}]).

@@

\begin{table*}
  \centering
  \begin{small}
  \begin{tabular}{|l|l|l|l|}
    \hline
    \multicolumn{4}{|c|}{\textbf{Affirmative forms}} \\
    \hline
    \textbf{English form} & \textbf{Example} & \textbf{Morphemes} & \textbf{Translation} \\
    \hline
    \multirow{2}{*}{Present simple} & \multirow{2}{*}{\emph{I play}} & \multirow{2}{*}{\emph{ойна} + <aorist> + <pers.>} & \emph{Мен ойнаймын} \\
                                    &                         &                                            & 'I play' \\
    \hline
    \multirow{2}{*}{Present continuous} & \multirow{2}{*}{\emph{I am playing}} & \emph{ойна} + <perf. part.>  & \emph{Мен ойнап жатырмын} \\
                                    &                         &   \emph{жатыр} + <pres.> + <pers.>     & 'I playing lie' \\
    \hline
    \multirow{2}{*}{Past continuous} & \multirow{2}{*}{\emph{I was playing}} & \emph{ойна} + <perf. part.>  & \emph{Мен ойнап отырдым} \\
                                    &                         & \emph{отыр} +  <past>  + <pers.>     & 'I playing sat' \\
    \hline
    \multirow{2}{*}{Obligation} & \multirow{2}{*}{I must go} & \emph{Мен}  + <gen.>  & \emph{Менің баруым керек} \\
                                    &                         & \emph{бар} + <ger.> +  <poss.\ p1 sg.> + \emph{керек}    & 'My going necessary [is]' \\
    \hline
    \multirow{2}{*}{\emph{Should}} & \multirow{2}{*}{\emph{I should go}} & Мен  + <gen.>  & \emph{Менің барғаным жөн} \\
                                    &                         & \emph{бар} + <past.ger.> +  <poss.\ p1 sg.> + \emph{жөн} & 'My having gone proper is' \\
    \hline
    \multicolumn{4}{|c|}{\textbf{Negative forms}} \\
    \hline
    \multirow{2}{*}{Present cont.} & \multirow{2}{*}{\emph{I am not playing}} & \emph{ойна} + <perf. part.>  & \emph{Мен ойнап жатқан жоқпын} \\
                                    &                         &   \emph{жатыр} + <pres.> + <pers.>     & 'I playing lying not' \\
    \hline
    \multirow{2}{*}{Inability} & \multirow{2}{*}{\emph{I can not play}} & \emph{ойна} + <impf. part.>  & \emph{Мен ойнай алмаймын} \\
                                    &                         &   \emph{ал} + <neg.> + <aor.> + <p1>     & 'I playing cannot' \\
    \hline
    \multirow{3}{*}{Present perf. cont. } & \multirow{3}{*}{\emph{You have not been playing}} & \emph{ойна} + <perf. part.>  & \emph{Сіз ойнап отырған емессі} \\
                                    &                         & +  \emph{отыр} + <past>  & 'You playing sat not'  \\
                                    &                         & + \emph{eмес} <cop.neg> + <p2> & \\
    \hline

  \end{tabular}
  \end{small}
  \caption{Examples of verb-phrase translations}
  \label{table:vps}
\end{table*}

\paragraph{Interchunk processing:}
